Gravitational Wave Detection:\\ On the performance of matched filtering and $\chi^2$ consistency testing in the presence of Gaussian noise, non–Gaussianly distributed noise, and transient glitches

### 2020-07-27 paper2.tex
Updates from Overleaf

\documentclass[reprint,
letterpaper,
 amsmath,amssymb,
 aps,
%floatfix,
]{revtex4-2}

@@

\begin{document}
\doublespacing
\title{Gravitational Wave Detection:\\
On the performance of the optimal matched filter in the presence of Gaussian noise, non–Gaussianly distributed noise, and transient glitches}
\author{Vinh Nguyen}
\email[Mentored by Dr. Eric Myers – Department of Physics and Astronomy, State University of New York at New Paltz]{}
\affiliation{2020 Pioneer Academics Program \\ Klein Oak High School, Spring, TX 77379}
\date{July 15, 2020}
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur mattis quam ligula. Suspendisse tempor ante felis, nec tincidunt risus pulvinar sed. Maecenas bibendum imperdiet odio, quis cursus lacus vulputate at. Donec pharetra cursus felis eget ultrices. Phasellus semper, lorem sit amet placerat interdum, ex purus semper nunc, nec accumsan nisl purus quis tellus. Etiam ac ornare augue. Integer ullamcorper malesuada ante, at mattis dui commodo sed. Donec commodo dictum turpis ut suscipit. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
\end{abstract}
\maketitle

\section{Introduction}
Ever since Einstein's first prediction of the existence of gravitational waves in 1915, gravitational wave science and detection have been a leading scientific field of interest. With the first confirmed detection of a binary black hole merger in 2015 (event GW150914), the prospects of gravitational wave science and its applications to other fields such as astronomy and fundamental physics have only become brighter. Gravitational wave detection, however, poses many challenges. This paper explores several challenges in the detection and data analysis of gravitational waves. Specifically, it studies the performance of the optimal matched filter in different noise conditions present in the Virgo and LIGO gravitational wave detectors.

\section{Background}
\subsection{Gravitational Waves}
In 1915, Einstein first published his theory of general relativity, describing space and time as one four-dimensional concept called \textit{``space-time''} \cite{maggiore_2008}, In this theory, he describes the force of gravity not as an inherent property that a massive body possesses, but rather as a byproduct of curved space-time. To put this more simply, we can say that a body with mass warps space-time, thereby creating a gravitational field surrounding it, affecting other bodies \cite{carlip}. To understand this idea, we can make use of an analogy: suppose that the fabric of space-time is a rubber sheet, and that any object with mass on it warps the surrounding area (FIG. 1.), hence creating a gravitational field. It follows from Einstein's Theory of Relativity that rapid changes in the curvature of space-time would produce ripples, called \textit{gravitational waves} (abbreviated as GW), that propagate outward at the speed of light  \cite{JSTORLIGO}.

\begin{figure}[h]
\includegraphics[width=8cm]{genral.png}
\caption{This figure shows a simplified analogy for Einstein's idea of space-time – which cannot be fully illustrated by a 2D picture. Adapted from \cite{mattson}.}
\centering
\end{figure}

\subsection{The Laser Interferometer Gravitational-Wave Observatory – LIGO}
LIGO (short for Laser Interferometer Gravitational-wave Observatory) is an NSF-funded, large scale scientific collaboration between the California Institute of Technology and the Massachusetts Institute of Technology designed to detect gravitational waves. As of June 2020, LIGO has two site locations – one in Hanford, Washington and the other in Livingston, Louisiana. Another important scientific collaboration is the Virgo project. The Virgo project is a European collaboration for gravitational wave detection funded by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale della Fisica Nucleare (INFN), the Dutch Nikhef in addition to the Polish and Hungarian institutes \cite{collaboration2019open}. Virgo has one detector located in Cascina, Italy. The two LIGO sites as well as the Virgo site in Italy each employs a type of detector called the Michelson Fabry-Perot Interferometer (interferometer for short) \cite{creighton_anderson_2011}. Virgo/LIGO's interferometers function by using four test masses (in this case, mirrors) in an L-shaped configuration, hung by wires near the ends and the vertex of the `L' as shown in FIG. 2. The test masses are configured so that the length $L_1 \approx L_2 = L$ (note that at the frequency of around $1$ Hz, the test masses can move freely horizontally) \cite{JSTORLIGO}.

\begin{figure}
\caption{A simplified diagram of the Michelson Fabry-Perot Interferometer. Adapted from \cite{JSTORLIGO}.}
\includegraphics[scale = .4]{Interferometer.png}
\centering
\end{figure} 
If there is a gravitational wave passing in between the test masses, the space between them will be stretched and compressed, squeezing one arm while stretching the other such that we can calculate the value of the arm length difference $\Delta L = L_1 - L_2$ \cite{teacherintro}. To measure this quantity $\Delta L$, the device uses a technique called laser interferometry \cite{LIGOArxiv}. Specifically, a powerful laser beam is shined at the beam splitter, thereby splitting the beam in half along the two arms. These beams then each enter a Fabry-Perot cavity where they are reflected back and forth between the mirrors, building up the laser light within the interferometer and increasing the distance travelled by the beam \cite{fabry} – increasing the interferometer's sensitivity. As the laser beams in the two arms exit the Fabry-Perot cavity and return to the beam splitter, the splitter recombines the two half-beams and gives off a portion of the combined beam to a photodiode – hence allowing scientists to analyze the interference patterns of the waves using optics principles, and thus determine the quantity $\Delta L$.

For gravitational waves passing through the test masses, the fractional difference in arm length $\frac{\Delta L}{L}$ is the strain $h$ (a quantity that measure the `strength' of gravitational waves) of the waves. Mathematically, it is a linear combination of its cross and plus polarizations, as follows \cite{JSTORLIGO}:
\begin{equation}
h(t) = \frac{\Delta L(t)}{L} = F_\times h_\times (t) + F_+ h_+ (t)
\end{equation}
\section{LIGO Data Analysis and Conventions}
\subsection{The Fourier Transform}
Virgo/LIGO signal is recorded as a function of time. However, to process this signal, we must usually transform the signal from its original time-domain into its frequency-domain through the use of a mathematical tool called the \textit{Fourier Transform}. In simplest terms, the Fourier Transform decomposes a function into its frequency content. Given a \textit{continuous} time-series function $F(t)$, its frequency-domain representation $\widetilde{F}(f)$ is given by \cite{dsptextbook}:
\begin{equation}
\widetilde{F}(f) = \int_{-\infty}^{\infty}F(t)e^{-2\pi\iu ft} dt
\end{equation}

@@

Since we are working with discrete-valued sequences, performing an DFT would give rise to spectral leakages. Thus, to minimize this, we always apply a windowing function to data in this paper whenever the use of the DFT is required (note that windowing does not decrease spectral leakage, but rather re-distributes the leakage effect so that it causes the least harm). In this paper, I also use a computer algorithm that implements the discrete Fourier Transform called the Fast Fourier Transform (FFT) which is implemented within the Python PyCBC library \cite{numpy}.

\subsection{Noise characteristics}
Being able to reach a strain sensitivity of $10^{-23}/\sqrt{\text{Hz}}$ \cite{sensitivity} means Virgo/LIGO's interferometers are extremely sensitive to any changes in the arm lengths – which can be caused by distorted space-time due to gravitational waves, or by vibrations that move the test masses. Thus, recorded signal $s(t)$ includes not only gravitational waves, but also unwanted vibrations from other sources as well \cite{ultimate}. Any unwanted noise, even on the smallest of scales, can still significantly interfere with data recorded by Virgo/LIGO's interferometers. Virgo/LIGO noise comes from many sources, but some examples include seismic noise due to motions of the Earth's surface, thermal noise at the molecular level affecting the test masses, and quantum shot noise at high frequencies caused by the quantum nature of light \cite{blair_howell_ju_zhao_2012}.

@@

\subsection{Matched filtering}
The primary goal of Virgo/LIGO data analysis is to determine if there is gravitational wave signal buried in the noise. There are many ways that this can be accomplished; however, the most efficient and effective method is called \textit{matched filtering}, especially in cases where noise is both Gaussian and stationary \cite{helstrom_1975}. Matched filtering allows the recorded data stream to be searched for the presence of a known gravitational wave template with varying arrival time and phase offset \cite{surf}.  A matched filter requires three components: the recorded data stream $s(t)$, the \textit{power spectral density (PSD)} function $S_n(t)$ of the noise, and a known waveform template $h(t)$ of the gravitational wave. Given these quantities, the complex-valued output of the matched filter at time $t=t_0$ can be computed as follows \cite{findchirp}:
\begin{equation}
    z(t_0) = 4\int_0^\infty\frac{\widetilde{s}(f)\widetilde{h}^*(f)}{S_n(f)}e^{2\pi\iu ft_0}df
\end{equation}
where $\widetilde{h}^*(f)$ denotes the complex conjugate of the Fourier Transform of the template strain $h(t)$. Notice that the existence of $S_n(f)$ essentially down-weighs the frequencies where the noise is loud and does the opposite where the noise is weak \cite{ultimate}. Given the complex matched filter output $z(t)$, we can compute a test statistic value $\rho$ called the \textit{signal-to-noise ratio} as follows \cite{findchirp}: 
\begin{equation}
   \rho (t) = \frac{|z(t)|}{\sigma}
\end{equation}
where $\sigma$ is a normalization factor is computed from \cite{findchirp}
\begin{equation}
   \sigma^2 =  4\int_0^\infty\frac{|\widetilde{h}(f)|^2}{S_n(f)}df
\end{equation}

@@

The methods of this paper sought to evaluate the performance of the matched filtering algorithm implemented in the PyCBC software package. The evaluation focused on 3 aspects of the signal extraction output:
\begin{itemize}
    \item Performance in the presence of perfectly Gaussian noise with $\mu = 0$  and $\sigma = 1$ scaled down by a factor of $10^{-21}$.
    \item Performance in the presence of non-Gaussian noise – in this case, Laplacian noise with location $\mu = 0$ and scale parameter $b = 1,2,3$. Laplacian noise with  $\mu = 0$ and varying $b$ was chosen for the purposes of this study due to its symmetric nature (similar to Gaussian noise) and varying levels of kurtosis. The higher the scale value $b$, the heavier the Laplacian noise distribution will be at the tails (note that heavier tails correspond to higher kurtoses). These Laplace distributed noise models were all scaled down by a factor of $10^{-21}$. An illustration of these distributions are shown in FIG. 3.
    \item Performance in the presence of transient glitches. In this paper, I used five samples of known Virgo/LIGO glitches with two \textit{`Extremely Loud'} glitches, one \textit{Scattered Light} glitch, one \textit{Whistle} glitch, and one \textit{Power Line} glitch.

@@

\begin{figure}[h]
\caption{The different probability density functions $f_X(x)$ are shown below. Since $\int_{-\infty}^\infty f_X(x)dx = 1$, the lower the  functions' maximums are, the heavier their tails must be, leading to higher kurtoses.}
\includegraphics[scale = .54]{pdf graphs.png}
\centering
\end{figure}

@@

For visualization purposes, strain time-series data of the glitches and their Q-transformed spectrograms are shown from FIG. 4. to FIG. 8.
\begin{figure}[h]
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 1. The bottom figure shows a Q-transformed spectrogram around Glitch 1. (Data queried from \cite{collaboration2019open})}
\includegraphics[scale = .6]{loud 69 graphics.png}
\centering
\end{figure} 
\begin{figure}[h]
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 2. The bottom figure shows a Q-transformed spectrogram around Glitch 2. (Data queried from \cite{collaboration2019open})}
\includegraphics[scale = .45]{loud 02 graphics.png}
\centering
\end{figure} 
\begin{figure}[h]
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 3. The bottom figure shows a Q-transformed spectrogram around Glitch 3. (Data queried from \cite{collaboration2019open})}
\includegraphics[scale = .75]{Scattered Light Graphics.png}
\centering
\end{figure} 
\begin{figure}[h]
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 4. The bottom figure shows a Q-transformed spectrogram around Glitch 4. (Data queried from \cite{collaboration2019open})}
\includegraphics[scale = .6]{power line graphics.png}
\centering
\end{figure} 
\begin{figure}[h]
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 5. The bottom figure shows a Q-transformed spectrogram around Glitch 5. (Data queried from \cite{collaboration2019open})}
\includegraphics[scale = .65]{whistle graphics.png}
\centering
\end{figure}

@@

I then proceeded to zero-pad the templates so that their lengths match the length of the 50-second stretch of noise. Afterwards, I `rolled' the templates so that they start at time $t=35$s in the 50-second stretch of data. In the case of Gaussian noise, the pure noise was generated via the Python NumPy package's \code{random.normal} method then scaled down both by a factor of $10^{-21}$ \cite{numpy}. In the case of Laplacian noise, the \code{random.laplace} method was used. It was then scaled down accordingly similar to the Gaussian case. As for the 5 samples of glitches, raw LIGO data of 25 seconds before and after the glitches' time of peak amplitude was queried directly from GWOSC (the code for this can be accessed at \url{https://tinyurl.com/yaedmpsx}) to be used as the noise (with glitch). GWOSC (short for Gravitational Wave Open Science Center) is a collaborative effort between the LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration to provide the public with gravitational wave data from the two LIGO facilities (Hanford, WA and Livingston, LA) as well the Virgo Observatory in Cascina, Italy. Note that these samples of glitches also contain background noise that may or may not be Gaussian and stationary in addition to the glitches themselves. Removing background noise from these samples and extracting these glitches require complex methods that were not attainable within the time frame of the writing of this paper.

@@

\begin{figure}[h]
\caption{This figure at the top shows the template waveform strain time-series of a binary black hole merger with masses 30 $M$\textsubscript{\(\odot\)} and 23 $M$\textsubscript{\(\odot\)} (Template 1). The figure at the bottom shows the template waveform strain time-series of a binary black hole merger with both black holes of mass 7 $M$\textsubscript{\(\odot\)} (Template 2). These waveforms were generated with PyCBC \cite{pycbc}.}
\includegraphics[scale = 1]{templates.png}
\centering
\end{figure}

@@

\begin{figure}[h]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from perfectly Gaussian noise scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 20.0$}
\includegraphics[scale = .33]{gaussian -21 template 1.png}
\centering
\end{figure}

@@

\begin{figure}[h!]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from perfectly Laplacian noise with $b=2$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 6.1$.}
\includegraphics[scale = .33]{laplacian b=2 template 1.png}
\centering
\end{figure} 
In the case of Laplacian noise with $b=2$ (FIG. 12.), the detection result was also accurate (i.e. there is a clear, thin, and distinct peak at time $t=35.0$). However, in this case, the SNR $\rho(t)$ is much lower, reaching only a value of $6.1$. Furthermore, it can be seen that the SNR `noise floor' is also relatively consistent with the Gaussian noise case and the $b=1$ Laplacian noise case, fluctuating around $\rho = 3.0$, though many times reaching $\rho = 4.0$.

@@

\begin{figure}[H]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from perfectly Laplacian noise with $b=3$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 21.7$s with SNR $\rho = 4.6$.}
\includegraphics[scale = .33]{laplacian b=3 template 1.png}
\centering
\end{figure}

@@

\begin{figure}[h]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from Glitch 1. A signal peak was detected at time $t = 17.7$s with SNR $\rho = 305.0$.}
\includegraphics[scale = .33]{glitch loud 69 template 1.png}
\centering
\end{figure}

\begin{figure}[h]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from Glitch 2. A signal peak was detected at time $t = 17.7$s with SNR $\rho = 294.6$.}
\includegraphics[scale = .33]{glitch loud 02 template 1.png}
\centering
\end{figure}

@@

\begin{figure}[h]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from Glitch 3. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 20.9$.}
\includegraphics[scale = .33]{glitch 3 template 1.png}
\centering
\end{figure}
\begin{figure}[h]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from Glitch 4. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 21.8$.}
\includegraphics[scale = .33]{glitch 4 template 1.png}
\centering
\end{figure}
\begin{figure}[h]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from Glitch 5. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 20.9$.}
\includegraphics[scale = .33]{glitch 5 template 1.png}
\centering
\end{figure}

>>>>

\documentclass[preprint,
letterpaper,
 amsmath,amssymb,
 aps,
%floatfix,
]{revtex4-2}

@@

\begin{document}
\doublespacing
\title{Gravitational Wave Detection:\\
On the performance of matched filtering and $\chi^2$ consistency testing in the presence of Gaussian noise, non–Gaussianly distributed noise, and transient glitches}
\author{Vinh Nguyen}
\email[Mentored by Dr. Eric Myers – Department of Physics and Astronomy, State University of New York at New Paltz]{}
\affiliation{2020 Pioneer Academics Program \\ Klein Oak High School, Spring, TX 77379}
\date{July 15, 2020}
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur mattis quam ligula. Suspendisse tempor ante felis, nec tincidunt risus pulvinar sed. Maecenas bibendum imperdiet odio, quis cursus lacus vulputate at. Donec pharetra cursus felis eget ultrices. Phasellus semper, lorem sit amet placerat interdum, ex purus semper nunc, nec accumsan nisl purus quis tellus. Etiam ac ornare augue. Integer ullamcorper malesuada ante, at mattis dui commodo sed. Donec commodo dictum turpis ut suscipit. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
\end{abstract}
\maketitle
\section{Introduction}
Ever since Einstein's first prediction of the existence of gravitational waves in 1915, gravitational wave science and detection have been a leading scientific field of interest. With the first confirmed detection of a binary black hole merger in 2015 (event GW150914), the prospects of gravitational wave science and its applications to other fields such as astronomy and fundamental physics have only become brighter. Gravitational wave detection, however, poses many challenges. This paper explores several challenges in the detection and data analysis of gravitational waves. Specifically, it studies the performance of the optimal matched filter in different noise conditions present in the Virgo and LIGO gravitational wave detectors.

This paper uses LIGO data queried from the Gravitational Wave Open Science Center (GWOSC) is a collaborative effort between the LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration to provide the public with gravitational wave data from the two LIGO facilities (Hanford, WA and Livingston, LA) as well the Virgo Observatory in Cascina, Italy

\section{Background}
\subsection{Gravitational waves and the Laser Interferometer Gravitational-Wave Observatory (LIGO)}
In 1915, Einstein first published his theory of general relativity, describing space and time as one four-dimensional concept called \textit{``space-time''} \cite{maggiore_2008}, In this theory, he describes the force of gravity not as an inherent property that a massive body possesses, but rather as a byproduct of curved space-time. To put this more simply, we can say that a body with mass warps space-time, thereby creating a gravitational field surrounding it, affecting other bodies \cite{carlip}. To understand this idea, we can make use of an analogy: suppose that the fabric of space-time is a rubber sheet, and that any object with mass on it warps the surrounding area (FIG. 1.), hence creating a gravitational field. It follows from Einstein's Theory of Relativity that rapid changes in the curvature of space-time would produce ripples, called \textit{gravitational waves} (abbreviated as GW), that propagate outward at the speed of light  \cite{JSTORLIGO}.

\begin{figure}[t]
\includegraphics[width=8cm]{genral.png}
\caption{This figure shows a simplified analogy for Einstein's idea of space-time – which cannot be fully illustrated by a 2D picture. Adapted from \cite{mattson}.}
\centering
\end{figure}

LIGO (short for Laser Interferometer Gravitational-wave Observatory) is an NSF-funded, large scale scientific collaboration between the California Institute of Technology and the Massachusetts Institute of Technology designed to detect gravitational waves. As of June 2020, LIGO has two site locations – one in Hanford, Washington and the other in Livingston, Louisiana. Another important scientific collaboration is the Virgo project. The Virgo project is a European collaboration for gravitational wave detection funded by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale della Fisica Nucleare (INFN), the Dutch Nikhef in addition to the Polish and Hungarian institutes \cite{collaboration2019open}. Virgo has one detector located in Cascina, Italy. The two LIGO sites as well as the Virgo site in Italy each employs a type of detector called the Michelson Fabry-Perot Interferometer (interferometer for short) \cite{creighton_anderson_2011}. Virgo/LIGO's interferometers function by using four test masses (in this case, mirrors) in an L-shaped configuration, hung by wires near the ends and the vertex of the `L' as shown in FIG. 2. The test masses are configured so that the length $L_1 \approx L_2 = L$ (note that at the frequency of around $1$ Hz, the test masses can move freely horizontally) \cite{JSTORLIGO}.

\begin{figure}
\includegraphics[scale = .4]{Interferometer.png}
\caption{A simplified diagram of the Michelson Fabry-Perot Interferometer. Adapted from \cite{JSTORLIGO}.}
\centering
\end{figure} 
If there is a gravitational wave passing in between the test masses, the space between them will be stretched and compressed, squeezing one arm while stretching the other such that we can calculate the value of the arm length difference $\Delta L = L_1 - L_2$ \cite{teacherintro}. To measure this quantity $\Delta L$, the device uses a technique called laser interferometry \cite{LIGOArxiv}. Specifically, a powerful laser beam is shined at the beam splitter, thereby splitting the beam in half along the two arms. These beams then each enter a Fabry-Perot cavity where they are reflected back and forth between the mirrors, building up the laser light within the interferometer and increasing the distance travelled by the beam \cite{fabry} – increasing the interferometer's sensitivity. As the laser beams in the two arms exit the Fabry-Perot cavity and return to the beam splitter, the splitter recombines the two half-beams and gives off a portion of the combined beam to a photodiode – hence allowing scientists to analyze the interference patterns of the waves using optics principles, and thus determine the quantity $\Delta L$.

For gravitational waves passing through the test masses, the fractional difference in arm length $\frac{\Delta L}{L}$ is the strain $h$ (a quantity that measure the `strength' of gravitational waves) of the waves. Mathematically, the strain $h(t) = \frac{\Delta L(t)}{L}$ can also be expressed as a linear combination of its cross and plus polarizations, as follows \cite{JSTORLIGO}:
\begin{equation}
h(t) = F_\times h_\times (t) + F_+ h_+ (t)
\end{equation}

\subsection{LIGO Data Analysis and Conventions}
\subsubsection{The Fourier Transform}
The signal from Virgo/LIGO is recorded as a function of time. However, to process this signal, we must usually transform the signal from its original time-domain into its frequency-domain through the use of a mathematical tool called the \textit{Fourier Transform}. In simplest terms, the Fourier Transform decomposes a function into its frequency content. Given a \textit{continuous} time-series function $F(t)$, its frequency-domain representation $\widetilde{F}(f)$ is given by \cite{dsptextbook}:
\begin{equation}
\widetilde{F}(f) = \int_{-\infty}^{\infty}F(t)e^{-2\pi\iu ft} dt
\end{equation}

@@

Since we are working with discrete-valued sequences, performing a DFT would give rise to spectral leakages. Thus, to minimize this, we always apply a windowing function to data in this paper whenever the use of the DFT is required (note that windowing does not decrease spectral leakage, but rather re-distributes the leakage effect so that it causes the least harm). In this paper, I also use a computer algorithm that implements the discrete Fourier Transform called the Fast Fourier Transform (FFT) which is implemented within the Python PyCBC library \cite{numpy}.

\subsubsection{Noise characteristics}
Being able to reach a strain sensitivity of $10^{-23}/\sqrt{\text{Hz}}$ \cite{sensitivity} means Virgo/LIGO's interferometers are extremely sensitive to any changes in the arm lengths – which can be caused by distorted space-time due to gravitational waves, or by vibrations that move the test masses. Thus, recorded signal $s(t)$ includes not only gravitational waves, but also unwanted vibrations from other sources as well \cite{ultimate}. Any unwanted noise, even on the smallest of scales, can still significantly interfere with data recorded by Virgo/LIGO's interferometers. Virgo/LIGO noise comes from many sources, but some examples include seismic noise due to motions of the Earth's surface, thermal noise at the molecular level affecting the test masses, and quantum shot noise at high frequencies caused by the quantum nature of light \cite{blair_howell_ju_zhao_2012}.

@@

\subsubsection{Matched filtering}
The primary goal of Virgo/LIGO data analysis is to determine if there is gravitational wave signal buried in the noise. There are many ways that this can be accomplished; however, the most efficient and effective method is called \textit{matched filtering}, especially in cases where noise is both Gaussian and stationary \cite{helstrom_1975}. Matched filtering allows the recorded data stream to be searched for the presence of a known gravitational wave template with varying arrival time and phase offset \cite{surf}.  A matched filter requires three components: the recorded data stream $s(t)$, the \textit{power spectral density (PSD)} function $S_n(t)$ of the noise, and a known waveform template $h(t)$ of the gravitational wave. Given these quantities, the complex-valued output of the matched filter at time $t=t_0$ can be computed as follows \cite{findchirp}:
\begin{equation}
    z(t_0) = 4\int_0^\infty\frac{\widetilde{s}(f)\widetilde{h}^*(f)}{S_n(f)}e^{2\pi\iu ft_0}df
\end{equation}
where $\widetilde{h}^*(f)$ denotes the complex conjugate of the Fourier Transform of the template strain $h(t)$. Notice that the existence of $S_n(f)$ essentially down-weighs the frequencies where the noise is loud and does the opposite where the noise is weak \cite{ultimate}. Given the complex matched filter output $z(t)$, we can compute a test statistic value $\rho$ called the \textit{signal-to-noise ratio} as follows \cite{findchirp}: 
\begin{equation}
   \rho (t) = \frac{|z(t)|}{\sigma}
\end{equation}
where $\sigma$ is a normalization factor is computed from \cite{findchirp}
\begin{equation}
   \sigma^2 =  4\int_0^\infty\frac{|\widetilde{h}(f)|^2}{S_n(f)}df
\end{equation}

@@

This paper sought to evaluate the performance of the matched filtering algorithm implemented in the PyCBC software package. The evaluation focused on 3 aspects of the signal extraction output:
\begin{itemize}
    \item Performance in the presence of perfectly Gaussian noise with $\mu = 0$  and $\sigma = 1$ scaled down by a factor of $10^{-21}$.
    \item Performance in the presence of non-Gaussian noise – in this case, Laplacian noise with location $\mu = 0$ and scale parameter $b = 1,2,3$. Laplacian noise with  $\mu = 0$ and varying $b$ was chosen for the purposes of this study due to its symmetric nature (similar to Gaussian noise) and varying levels of kurtosis. The higher the scale value $b$, the heavier the Laplacian noise distribution will be at the tails (note that heavier tails correspond to higher kurtoses). These Laplace distributed noise models were all scaled down by a factor of $10^{-21}$. An illustration of these distributions are shown in FIG. 3.
    \item Performance in the presence of transient glitches. In this paper, I used five samples of known Virgo/LIGO glitches with two \textit{`Extremely Loud'} glitches, one \textit{Scattered Light} glitch, one \textit{Whistle} glitch, and one \textit{Power Line} glitch.

@@

\begin{figure}[t]
\caption{The different probability density functions $f_X(x)$ are shown below. Since $\int_{-\infty}^\infty f_X(x)dx = 1$, the lower the  functions' maximums are, the heavier their tails must be, leading to higher kurtoses.}
\includegraphics[scale = .54]{pdf graphs.png}
\centering
\end{figure}

@@

For visualization purposes, strain time-series data of the glitches and their Q-transformed spectrograms are shown from FIG. 4. to FIG. 8.
\begin{figure}[t]
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 1. The bottom figure shows a Q-transformed spectrogram around Glitch 1. (Data queried from \cite{collaboration2019open})}
\includegraphics[scale = .6]{loud 69 graphics.png}
\centering
\end{figure} 
\begin{figure}[t]
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 2. The bottom figure shows a Q-transformed spectrogram around Glitch 2. (Data queried from \cite{collaboration2019open})}
\includegraphics[scale = .45]{loud 02 graphics.png}
\centering
\end{figure} 
\begin{figure}[t]
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 3. The bottom figure shows a Q-transformed spectrogram around Glitch 3. (Data queried from \cite{collaboration2019open})}
\includegraphics[scale = .75]{Scattered Light Graphics.png}
\centering
\end{figure} 
\begin{figure}[t]
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 4. The bottom figure shows a Q-transformed spectrogram around Glitch 4. (Data queried from \cite{collaboration2019open})}
\includegraphics[scale = .6]{power line graphics.png}
\centering
\end{figure} 
\begin{figure}[t]
\caption{The top figure shows the strain time-series of 50 seconds around Glitch 5. The bottom figure shows a Q-transformed spectrogram around Glitch 5. (Data queried from \cite{collaboration2019open})}
\includegraphics[scale = .65]{whistle graphics.png}
\centering
\end{figure}

@@

I then proceeded to zero-pad the templates so that their lengths match the length of the 50-second stretch of noise. Afterwards, I `rolled' the templates so that they start at time $t=35$s in the 50-second stretch of data. In the case of Gaussian noise, the pure noise was generated via the Python NumPy package's \code{random.normal} method then scaled down both by a factor of $10^{-21}$ \cite{numpy}. In the case of Laplacian noise, the \code{random.laplace} method was used. It was then scaled down accordingly similar to the Gaussian case. As for the 5 samples of glitches, raw LIGO data of 25 seconds before and after the glitches' time of peak amplitude was queried directly from GWOSC (the code
for this can be accessed at \footnote{V.Nguyen, Google Colab notebook for signal matched filtering and chi-squared consistency testing, \href{https://colab.research.google.com/drive/10q3SdhNjWyoZaVPdLqyoCEsSovplvMrK?authuser=2#scrollTo=7cafQoPiH5Sy}{(2020)}. URL: \url{https://colab.research.google.com/drive/10q3SdhNjWyoZaVPdLqyoCEsSovplvMrK?authuser=2#scrollTo=7cafQoPiH5Sy}}) to be used as the noise (with glitch). Note that these samples of glitches also contain background noise that may or may not be Gaussian and stationary in addition to the glitches themselves. Removing background noise from these samples and extracting these glitches require complex methods that were not attainable within the time frame of the writing of this paper.

@@

\begin{figure}[t]
\caption{This figure at the top shows the template waveform strain time-series of a binary black hole merger with masses 30 $M$\textsubscript{\(\odot\)} and 23 $M$\textsubscript{\(\odot\)} (Template 1). The figure at the bottom shows the template waveform strain time-series of a binary black hole merger with both black holes of mass 7 $M$\textsubscript{\(\odot\)} (Template 2). These waveforms were generated with PyCBC \cite{pycbc}.}
\includegraphics[scale = 1]{templates.png}
\centering
\end{figure}

@@

\begin{figure}[t]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from perfectly Gaussian noise scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 20.0$}
\includegraphics[scale = .33]{gaussian -21 template 1.png}
\centering
\end{figure}

@@

\begin{figure}[t]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from perfectly Laplacian noise with $b=2$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 6.1$.}
\includegraphics[scale = .33]{laplacian b=2 template 1.png}
\centering
\end{figure} 
In the case of Laplacian noise with $b=2$ (FIG. 12.), the detection result was also accurate (i.e. there is a clear, thin, and distinct peak at time $t=35.0$). However, in this case, the SNR $\rho(t)$ is much lower, reaching only a value of $6.1$. Furthermore, it can be seen that the SNR `noise floor' is also relatively consistent with the Gaussian noise case and the $b=1$ Laplacian noise case, fluctuating around $\rho = 3.0$, though many times reaching $\rho = 4.0$.

@@

\begin{figure}[t]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from perfectly Laplacian noise with $b=3$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 21.7$s with SNR $\rho = 4.6$.}
\includegraphics[scale = .33]{laplacian b=3 template 1.png}
\centering
\end{figure}

@@

\begin{figure}[t]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from Glitch 1. A signal peak was detected at time $t = 17.7$s with SNR $\rho = 305.0$.}
\includegraphics[scale = .33]{glitch loud 69 template 1.png}
\centering
\end{figure}

\begin{figure}[t]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from Glitch 2. A signal peak was detected at time $t = 17.7$s with SNR $\rho = 294.6$.}
\includegraphics[scale = .33]{glitch loud 02 template 1.png}
\centering
\end{figure}

@@

\begin{figure}[t]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from Glitch 3. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 20.9$.}
\includegraphics[scale = .33]{glitch 3 template 1.png}
\centering
\end{figure}
\begin{figure}[t]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from Glitch 4. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 21.8$.}
\includegraphics[scale = .33]{glitch 4 template 1.png}
\centering
\end{figure}
\begin{figure}[t]
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after a matched filter was applied to recover Template 1 from Glitch 5. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 20.9$.}
\includegraphics[scale = .33]{glitch 5 template 1.png}
\centering
\end{figure}
